---
title: "Reproducible Research Assignment 2"
author: "Project_H"
date: "24 April 2016"
output: 
  html_document:
    keep_md: yes
---

# NOAA Storm Database Snapshot Analysis: Synopsis

This is an analysis of a snapshot of the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database, from 18 April 1950 to 29 November 2011. This analysis is the second assignment in the Reproducible Research class. From the data, we determined that the floods had the costliest economic consequences, while tornados caused the most fatalities and injuries.

## Software Environment, Data Files, and External References:
This analysis was built using:  
* R 3.2.4  
* data.table 1.9.6  
* readr 0.2.2

The data used in this analysis is a snapshot located at:  
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Interpretation of the dataset variables, specifically the PROPDMGEXP and CROPDMGEXP fields, was informed by 'How To Handle Exponent Value of PROPDMGEXP and CROPDMGEXP' located at:  
https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html

```{r required_packages}
require(data.table)
require(readr)
require(lattice)
```

## Data Processing

### Loading Data
Unfortunately, the data file crashes data.table 1.9.6's fread function. readr's read_csv is used to load the data, as it is an order of magnitude faster than the base read.csv. 
```{r load_data}
dataset <- data.table(read_csv('../repdata-data-StormData.csv.bz2'))
```

### Pre-Processing before Analysis
The dataset needs some pre-processing before it may be used for analysis.

#### PROPDMGEXP and CROPDMGEXP Fields
The PROPDMGEXP and CROPDMGEXP fields are converted to a uniform number base (in dollars), as informed by 'How To Handle Exponent Value of PROPDMGEXP and CROPDMGEXP', as the value levels are not explained in NOAA's storm database documentation.

```{r dmg_exp_fields}
dataset$PROPDMGEXP <- as.factor(dataset$PROPDMGEXP)
levels(dataset$PROPDMGEXP)

# Convert PROPDMG to dollars, based on the analysis at:
# https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html
dataset[PROPDMGEXP == '+', PropertyDamage := PROPDMG]
dataset[PROPDMGEXP %like% '\\d', PropertyDamage := PROPDMG * 10]
dataset[PROPDMGEXP %like% 'H|h', PropertyDamage := PROPDMG * 100]
dataset[PROPDMGEXP == 'K', PropertyDamage := PROPDMG * 1000]
dataset[PROPDMGEXP %like% 'M|m', PropertyDamage := PROPDMG * 10^6]
dataset[PROPDMGEXP == 'B', PropertyDamage := PROPDMG * 10^9]

# Convert CROPDMG to dollars:
dataset[CROPDMGEXP == '+', CropDamage := CROPDMG]
dataset[CROPDMGEXP %like% '\\d', CropDamage := CROPDMG * 10]
dataset[CROPDMGEXP %like% 'H|h', CropDamage := CROPDMG * 100]
dataset[CROPDMGEXP == 'K', CropDamage := CROPDMG * 1000]
dataset[CROPDMGEXP %like% 'M|m', CropDamage := CROPDMG * 10^6]
dataset[CROPDMGEXP == 'B', CropDamage := CROPDMG * 10^9]

# Sum CROPDMG and PROPDMG into TotalDamage. Not a simple addition formula because NAs will not properly sum. NA + 1 = NA, not 1.
dataset[, TotalDamage := rowSums(.SD, na.rm = TRUE), .SDcols = c('PropertyDamage','CropDamage')]

```

#### EVTYPE field

The dataset has huge number of unique 'event types': **`r length(unique(dataset$EVTYPE))`**

Many 'event types' describe the same type of event, and only vary in spelling, punctuation, and word order.

To reduce the variation in event types, we're going to:  
* extract a summary table of event types with totals for Fatalities, Injuries, PropertyDamage, and CropDamage  
* standardize the various event type names, where possible 
* melt and re-cast the summary table which results in a smaller table with a reduced number of event types 

This is not a perfect method of reducing Event Type variation, as some 700+ event types still exist after name standardization. In part this is due to Event Types that include multiple phenomenona which may not be equivalent. For example, 'BEACH EROSION/COASTAL FLOOD' may not be equivalent to 'COASTAL FLOOD AND EROSION'.

```{r evtype_standardization}
# Extract a summary of storm data events, a count of each EventType, and merge the two tables. EVCOUNT is forced to double precision, to suppress downstream warning messages from melt.data.table.
# There is a bug in data.table 1.9.6. When merging the two tables keyed on the EVTYPE column, the merge function does not perfectly preserve the EVTYPE data. 
StormSum1 <- dataset[, lapply(.SD, sum, na.rm=TRUE), by=EVTYPE, 
             .SDcols=c('FATALITIES', 'INJURIES', 'PropertyDamage', 'CropDamage', 'TotalDamage')]
StormCount1 <- dataset[, .N, by = EVTYPE]
names(StormCount1)[2] <- 'EVCOUNT'
StormCount1$EVCOUNT <- as.double(StormCount1$EVCOUNT)
StormSum2 <- StormSum1[StormCount1, on='EVTYPE']

# Convert all EventTypes to upper case to reduce variation in Event Types
StormSum2[,EVTYPE := toupper(EVTYPE)]

# Condense multiple spaces between words to one space between words
StormSum2$EVTYPE <- gsub('\\s+', ' ', StormSum2$EVTYPE)

# Get rid of spelling, punctuation, word order variations. Double escapes for literals are required because the regular expression is passed through data.table to grepl
StormSum2[EVTYPE == 'AVALANCE',  EVTYPE := 'AVALANCHE']
StormSum2[EVTYPE == 'BEACH EROSIN',  EVTYPE := 'BEACH EROSION']
StormSum2[EVTYPE %like% '^BLOWING SNOW[ &/-]*EXTREME', EVTYPE := 'BLOWING SNOW EXTREME WIND CHILL']
StormSum2[EVTYPE %like% 'BLOW-OUT TIDE[S]*', EVTYPE := 'BLOWOUT TIDE']
StormSum2[EVTYPE %like% '^BRUSH FIRE[S]*', EVTYPE := 'BRUSH FIRE']
StormSum2[EVTYPE %like% '^COASTAL[ ]*FLOOD(ING)*$', EVTYPE := 'COASTAL FLOOD']
StormSum2[EVTYPE %like% 'C(OA)*ST(A)*L[ ]*FLOODING/EROSION', EVTYPE := 'COASTAL FLOOD AND EROSION']
StormSum2[EVTYPE == 'EROSION/CSTL FLOOD', EVTYPE := 'COASTAL FLOOD AND EROSION']
StormSum2[EVTYPE %like% '^COASTAL[ ]*STORM', EVTYPE := 'COASTAL STORM']
StormSum2[EVTYPE %like% 'COLD AIR FUNNEL[S]*', EVTYPE := 'COLD AIR FUNNEL']
StormSum2[EVTYPE == 'COLD TEMPERATURES', EVTYPE := 'COLD TEMPERATURE']
StormSum2[EVTYPE %like% 'DUST DEV[EI]L$', EVTYPE := 'DUST DEVIL']
StormSum2[EVTYPE %like% 'EXTREME WIND[ ]*CHILL[S]*$', EVTYPE := 'EXTREME WIND CHILL']
StormSum2[EVTYPE %like% 'FLOO*D[/]*FLASH', EVTYPE := 'FLASH FLOOD']
StormSum2[EVTYPE %like% '^FLASH.*?FLOO[O]*D(S|ING)*$', EVTYPE := 'FLASH FLOOD']
StormSum2[EVTYPE %like% '^FLASH FLOOD/$', EVTYPE := 'FLASH FLOOD']
StormSum2[EVTYPE %like% 'FLOOD[ /]FLASH', EVTYPE := 'FLASH FLOOD']
StormSum2[EVTYPE %like% '^FUNNEL CLOUD[\\.S]*$', EVTYPE := 'FUNNEL CLOUD']
StormSum2[EVTYPE %like% 'FROST[/\\]FREEZE', EVTYPE := 'FROST FREEZE']
StormSum2[EVTYPE %like% '^GRADIENT WIND[S]*$', EVTYPE := 'GRADIENT WIND']
StormSum2[EVTYPE %like% '^GUSTY WIND[S]*$', EVTYPE := 'GUSTY WIND']
StormSum2[EVTYPE %like% '^HAIL.*?( |0\\.|0)75$', EVTYPE := 'HAIL 075']
StormSum2[EVTYPE %like% '^HAIL\\(0\\.75', EVTYPE := 'HAIL 075']
StormSum2[EVTYPE %like% '^HAIL 80', EVTYPE := 'HAIL 080']
StormSum2[EVTYPE %like% '^HAIL.*?88$', EVTYPE := 'HAIL 088']
StormSum2[EVTYPE %like% '^HAIL 1[\\.]*00', EVTYPE := 'HAIL 100']
StormSum2[EVTYPE %like% '^HAIL 1[\\.]*75[\\)]*$', EVTYPE := 'HAIL 175']
StormSum2[EVTYPE %like% '^HAILSTORM[S]*$', EVTYPE := 'HAILSTORM']
StormSum2[EVTYPE %like% '^HAIL/WIND[S]*$', EVTYPE := 'HAIL WIND']
StormSum2[EVTYPE %like% '^HEAT WAVE[S]*$', EVTYPE := 'HEAT WAVE']
StormSum2[EVTYPE %like% '^HEAVY .*?BLOWING.*?SNOW', EVTYPE := 'HEAVY BLOWING SNOW']
StormSum2[EVTYPE %like% '^HEAVY RAIN[S]*$', EVTYPE := 'HEAVY RAIN']
StormSum2[EVTYPE %like% '^HEAVY SHOWER[S]*$', EVTYPE := 'HEAVY SHOWER']
StormSum2[EVTYPE %like% 'HEAVY SNOW[ /]*FREEZING RAIN', EVTYPE := 'HEAVY SNOW FREEZING RAIN']
StormSum2[EVTYPE %like% '^HEAVY SNOW.*?ICE STORM', EVTYPE := 'HEAVY SNOW ICE STORM']
StormSum2[EVTYPE %like% '^HEAVY SNOW[- ]SQUALLS$', EVTYPE := 'HEAVY SNOW SQUALL']
StormSum2[EVTYPE %like% '^HIGH WIND[S/]*$', EVTYPE := 'HIGH WINDS']
StormSum2[EVTYPE %like% '^LAKE[ -]*EFFECT SNOW', EVTYPE := 'LAKE EFFECT SNOW']
StormSum2[EVTYPE %like% '^LANDSLIDE[S]*$', EVTYPE := 'LANDSLIDE']
StormSum2[EVTYPE %like% '^LI.*?ING[\\.]*$', EVTYPE := 'LIGHTNING']
StormSum2[EVTYPE %like% '^LIGHTNING( AND |/)HEAVY RAIN', EVTYPE := 'LIGHTNING HEAVY RAIN']
StormSum2[EVTYPE %like% '^MARINE (TSTM|THUNDERSTORM) WIND', EVTYPE := 'MARINE THUNDERSTORM WIND']
StormSum2[EVTYPE %like% 'MUD[ ]*SLIDE[S]*$', EVTYPE := 'MUDSLIDE']
StormSum2[EVTYPE %like% '^RIP CURRENT[S]*$', EVTYPE := 'RIP CURRENT']
StormSum2[EVTYPE %like% '^RIP CURRENTS[ /]HEAVY SURF', EVTYPE := 'RIP CURRENT HEAVY SURF']
StormSum2[EVTYPE %like% '^FREEZING RAIN( AND |/)SLEET', EVTYPE := 'SLEET FREEZING RAIN']
StormSum2[EVTYPE %like% '^SLEET[ &/]*FREEZING RAIN', EVTYPE := 'SLEET FREEZING RAIN']
StormSum2[EVTYPE %like% '^SNOW[ /]*FREEZING RAIN', EVTYPE := 'SNOW FREEZING RAIN']
StormSum2[EVTYPE %like% '^FREEZING RAIN( AND |/)SNOW', EVTYPE := 'SNOW FREEZING RAIN']
StormSum2[EVTYPE %like% '^SNOW.*?ICE$', EVTYPE := 'SNOW ICE']
StormSum2[EVTYPE %like% '^SNOW.*?ICE STORM', EVTYPE := 'SNOW ICE STORM']
StormSum2[EVTYPE %like% '^SNOW SQUALL[S]*', EVTYPE := 'SNOW SQUALL']
StormSum2[EVTYPE %like% '^STRONG WIND[S]*$', EVTYPE := 'STRONG WIND']
StormSum2[EVTYPE %like% '^TIDAL FLOOD(ING)*$', EVTYPE := 'TIDAL FLOOD']
StormSum2[EVTYPE == 'TORNDAO',  EVTYPE := 'TORNADO']
StormSum2[EVTYPE %like% '^TORNADO(ES|S)$', EVTYPE := 'TORNADO']
StormSum2[EVTYPE %like% '^WATER *SPOUT[-/S]*$', EVTYPE := 'WATERSPOUT']
StormSum2[EVTYPE == 'WAYTERSPOUT',  EVTYPE := 'WATERSPOUT']
StormSum2[EVTYPE %like% '^WATERSPOUT[ /-]*TORNADO$', EVTYPE := 'WATERSPOUT TORNADO']
StormSum2[EVTYPE == 'TORNADO/WATERSPOUT', EVTYPE := 'WATERSPOUT TORNADO']
StormSum2[EVTYPE == 'WET MICOBURST',  EVTYPE := 'WET MICROBURST']
StormSum2[EVTYPE %like% '^WILD.*?FIRES*$', EVTYPE := 'WILDFIRE']
StormSum2[EVTYPE %like% '^WI*NDS*$', EVTYPE := 'WIND']
StormSum2[EVTYPE %like% 'WINTER WEATHER[ /]MIX', EVTYPE := 'WINTRY MIX']
StormSum2[EVTYPE %like% 'WINT(E?RY?) MIX', EVTYPE := 'WINTRY MIX']
StormSum2[EVTYPE %like% 'WINTER STORM[ /]HIGH WINDS*', EVTYPE := 'WINTER STORM HIGH WINDS']
StormSum2[EVTYPE == 'WINTER STORMS', EVTYPE := 'WINTER STORM']
StormSum2[EVTYPE %like% '^URBAN FLOOD[S|ING]*$', EVTYPE := 'URBAN FLOODING']
StormSum2[EVTYPE %like% '^URBAN.*?[SMALL|SML].*?STREAM.*?[FLOOD|FLD]*', EVTYPE := 'URBAN SMALL STREAM FLOODING']
StormSum2[EVTYPE %like% 'SMALL STREAM.*?URBAN FLOOD', EVTYPE := 'URBAN SMALL STREAM FLOODING']
StormSum2[EVTYPE == 'UNSEASONABLE COLD',  EVTYPE := 'UNSEASONABLY COLD']
StormSum2[EVTYPE %like% '^UNSEASONABLY WARM[ &/]*WET', EVTYPE := 'UNSEASONABLY WARM WET']

# Goal: THUNDERSTORM WIND. Lots of different patterns in Grep.
# Starting with TH and ending in ND or DS or DSS or ND. or NDS or INS.
# Metacharacter: . = any character
#              : * = Match the preceding element 0 or more times
#              : ? = Match the preceding element 0 or 1 time.
StormSum2[EVTYPE %like% '^TH.*?(ND$|DS[S]*$|ND[S]*\\.$|INS)', EVTYPE := 'THUNDERSTORM WIND']
StormSum2[EVTYPE %like% 'THUNDERSTORMW', EVTYPE := 'THUNDERSTORM WIND']
StormSum2[EVTYPE =='TUNDERSTORM WIND', EVTYPE := 'THUNDERSTORM WIND']

# Starting with 'TSTM W' and ending IN ND, NDS
StormSum2[EVTYPE %like% '^TSTM W.*?(ND$|DS$)', EVTYPE := 'THUNDERSTORM WIND']
StormSum2[EVTYPE %like% '^TSTMW', EVTYPE := 'THUNDERSTORM WIND']

# with numbers
StormSum2[EVTYPE %like% '^(TSTM|THUN).*?40', EVTYPE := 'THUNDERSTORM WIND 40']
StormSum2[EVTYPE %like% '^TSTM WIND.*?45', EVTYPE := 'THUNDERSTORM WIND 45']
StormSum2[EVTYPE %like% '^(TSTM|THUN).*?50', EVTYPE := 'THUNDERSTORM WIND 50']
StormSum2[EVTYPE %like% '^(TSTM|THUN).*?52', EVTYPE := 'THUNDERSTORM WIND 52']
StormSum2[EVTYPE %like% '^(TSTM|THUN).*?53', EVTYPE := 'THUNDERSTORM WIND 53']
StormSum2[EVTYPE %like% '^(TSTM|THUN).*?59', EVTYPE := 'THUNDERSTORM WIND 59']
StormSum2[EVTYPE %like% '^THUN.*?60', EVTYPE := 'THUNDERSTORM WIND 60']
StormSum2[EVTYPE %like% '^THUN.*?61', EVTYPE := 'THUNDERSTORM WIND 61']
StormSum2[EVTYPE %like% '^(TSTM|THUN).*?65', EVTYPE := 'THUNDERSTORM WIND 65']

# with hail
StormSum2[EVTYPE %like% 'THUNDERSTORM WIND.*?HAIL', EVTYPE := 'THUNDERSTORM WIND HAIL']
StormSum2[EVTYPE %like% '^TSTM WIND.*?HAIL', EVTYPE := 'THUNDERSTORM WIND HAIL']

# other
StormSum2[EVTYPE %like% 'THUNDERSTORM WIND.*?TREE[S]*', EVTYPE := 'THUNDERSTORM WIND TREES']


# Melt the summary to prepare for re-calculating the sums for each Event Type.
StormMelt <- melt.data.table(StormSum2, id.vars = 'EVTYPE', variable.name = 'DamageType')

# Redo the summary table, which essentially collapses duplicate rows created by correcting spelling and punctuation variations in EventType.
StormSum3 <- dcast(StormMelt, EVTYPE ~ DamageType, value.var = 'value', fun.aggregate = sum )

```

## Results

### Events with Greatest Economic Consequences

We will define 'Economic Consequences' as the sum of Property Damage and Crop Damage. We'll select and graph the Top 25 event types for Total Damage. 

```{r EconomicDamage}
# Select the Top 25 event types by Total Damage descending, and reduce the scale from dollars to Billions of Dollars.
EconDam25 <- head(StormSum3[order (-TotalDamage)], 25)
EconDam25[, TotalDamage := TotalDamage / 10^9]
EconDam25[, PropertyDamage := PropertyDamage / 10^9]
EconDam25[, CropDamage := CropDamage / 10^9]

# Even though the data table is sorted by Total Damage descending, the setting will not hold for the graph. This reorder is specifically done for the graph, so that barchart plots with greatest Total Damage values descending, instead of being graphed alphabetically by Event Type.
EconDam25$EVTYPE <- reorder(EconDam25$EVTYPE, EconDam25$TotalDamage) 

barchart(EVTYPE ~ PropertyDamage + CropDamage, data=EconDam25, horizontal = TRUE, stack = TRUE, 
         xlim = c(0,160), main = 'Economic Consequences of Storms: Top 25 Event Types',
         scales=list(cex=c(0.60,1), tick.number=8),
         auto.key=list(space='top', columns=2, cex=0.75),
         xlab=list(label='Total Damage in $ billions'))

# List the Top 5 (for brevity) Event Types for Total Damage.
EconDam25[1:5, .(EVTYPE, TotalDamage = prettyNum(TotalDamage, small.mark = '.', digits=4, width=5))]

```


Per the graph, the event type with the greatest economic consequence is: **`r EconDam25[,EVTYPE[1]]`**

The Event Type of second-great economic consequence is: **`r EconDam25[,EVTYPE[2]]`**

Most of the economic damage from sustained in the Top 25 event types is Property Damage. Floods cause the most economic damage, followed distantly by 'Hurricane/Typhoon' events.

Drought, Ice Storms, Extreme Cold, and Frost Freeze seem to be the only event types where Crop Damage is greater than Property Damage.

A potential issue is that these financial numbers may not have been indexed for inflation.


### Events with Greatest Impact on Population Health

There does not seem to be a standard single 'rating' of Population Health Impact based on fatalities and injuries. As a result, we will report on both separately.

```{r Fatalities}
Fatal25 <- head(StormSum3[order (-FATALITIES)], 25)

# Reorder by Event Type by fatalities for the graph, otherwise the graph will be presented by Event Type, in an alphabetical descending manner.
Fatal25$EVTYPE <- reorder(Fatal25$EVTYPE, Fatal25$FATALITIES) 

barchart(EVTYPE ~ FATALITIES, data=Fatal25, horizontal = TRUE, 
         main = 'Storm Fatalities: Top 25 Event Types',
         scales=list(cex=c(0.60,1), tick.number=14),
         xlab=list(label='Number of Fatalities'))

# List the Top 5 (for brevity) Event Types for fatalities.
Fatal25[1:5, .(EVTYPE, Fatalities = prettyNum(FATALITIES, big.mark = ','))]

```

The Event Type with the most fatalities is: **`r Fatal25[,EVTYPE[1]]`**

The Event Type with the second-most fatalities is: **`r Fatal25[,EVTYPE[2]]`**


```{r Injuries}
Injury25 <- head(StormSum3[order (-INJURIES)], 25)
Injury25$EVTYPE <- reorder(Injury25$EVTYPE, Injury25$INJURIES) 

barchart(EVTYPE ~ INJURIES, data=Injury25, horizontal = TRUE,  
         main = 'Storm Injuries: Top 25 Event Types',
         scales=list(cex=c(0.60,1), tick.number=10),
         xlab=list(label='Number of Injuries'))

# List the Top 5 (for brevity) Event Types for injuries.
Injury25[1:5, .(EVTYPE, Injuries = prettyNum(INJURIES, big.mark= ','))]

```

The Event Type with the most injuries is: **`r Injury25[,EVTYPE[1]]`**

The Event Type with the second-most injuries is: **`r Injury25[,EVTYPE[2]]`**
